---
---
title: "Learning Spatial Transcriptomics: Human and NHP Comparison"
author: "Template Creator"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    
This vignette has been made to compare expression of one or more tissues between humans and NHPs. 
---

```{r configuration, include=FALSE}
# ==============================================================================
# 0. CORE DEPENDENCIES & USER CONFIGURATION
# ==============================================================================

#When prompted about package compiling, select "No"
if (!requireNamespace("knitr", quietly = TRUE)) {
  install.packages("knitr", repos = "http://cran.us.r-project.org")
}
if (!requireNamespace("rmarkdown", quietly = TRUE)) {
  install.packages("rmarkdown", repos = "http://cran.us.r-project.org")
}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# ------------------------------------------------------------------------------
# GLOBAL DIRECTORIES & VARIABLES
# ------------------------------------------------------------------------------
# Automatically detect the user's home directory (e.g., C:\Users\rblair3)
USER_HOME <- Sys.getenv("USERPROFILE") 
if (USER_HOME == "") USER_HOME <- Sys.getenv("HOME") # Fallback for Mac/Linux

# Define the master working directory where data is loaded AND outputs are saved.
# This builds: C:\Users\[Username]\Box\LST\CosMX 6K NHP and Human TMA
BASE_DIR <- file.path(USER_HOME, "Box", "LST", "CosMX 6K NHP and Human TMA")

# Set global seed for reproducibility (e.g., UMAP generation, clustering)
GLOBAL_SEED <- 42

# Select datasets to load: c("NHP", "Human") or just c("Human")
DATASET_TARGETS <- c("NHP", "Human") 

# Define which column in your metadata contains the target variables (Case-insensitive)
SUBSET_COLUMN <- "Tissue" 

# Define target values (Leave as NULL to load all data)
TARGET_VALUES <- c("Liver")
```

#Library setup
```{r libraries, message=FALSE, warning=FALSE}
# ==============================================================================
# 1. ROBUST INSTALLER & LIBRARY SETUP
# ==============================================================================
set.seed(GLOBAL_SEED)
print(paste("✅ Global Seed Set to:", GLOBAL_SEED))

# The most robust way to handle dependencies in shared templates is 'pacman'.
# It automatically checks if a package exists, installs it if missing, and loads it.
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman", repos = "http://cran.us.r-project.org")
}

# 1A. LOAD STANDARD CRAN PACKAGES
pacman::p_load(
  Matrix,
  data.table,
  Seurat,
  ggplot2,
  patchwork,
  dplyr,
  readxl,
  remotes,
  clustree,
  ggrepel,
  harmony,
  R.utils
)

# 1B. LOAD CUSTOM GITHUB TOOLS (CosMx Scratch Space)
install_cosmx_tools <- function() {
  
  tools <- list("scPearsonPCA" = "_code/scPearsonPCA")
  missing_tools <- names(tools)[!sapply(names(tools), require, character.only = TRUE, quietly = TRUE)]
  
  if (length(missing_tools) == 0) {
    print("✅ Custom CosMx tools are already installed.")
    return(TRUE)
  }
  
  print(paste("⚠️ Missing custom tools:", paste(missing_tools, collapse=", "), "- Starting installer..."))

  # Force Fresh Download of the Nanostring Scratch Space Repository
  zip_dest <- file.path(tempdir(), "CosMx_Repo.zip")
  unzip_dest <- file.path(tempdir(), "CosMx_Unzipped")
  
  if (file.exists(zip_dest)) unlink(zip_dest, force = TRUE)
  if (dir.exists(unzip_dest)) unlink(unzip_dest, recursive = TRUE, force = TRUE)
  
  print("⬇️ Downloading CosMx Repository...")
  url <- "https://github.com/Nanostring-Biostats/CosMx-Analysis-Scratch-Space/archive/refs/heads/Main.zip"
  
  # FIX: Temporarily increase R's download timeout from 60 seconds to 1 hour
  old_timeout <- getOption("timeout")
  options(timeout = 3600)
  
  tryCatch({
    download.file(url, destfile = zip_dest, mode = "wb", method = "auto")
  }, error = function(e) {
    stop("❌ Download failed. Check internet connection or firewall settings.")
  })
  
  print("📦 Unzipping...")
  unzip(zip_dest, exdir = unzip_dest)
  
  # Dynamic Path Finding for extracted repo
  extracted_folders <- list.dirs(unzip_dest, full.names = TRUE, recursive = FALSE)
  repo_root <- extracted_folders[1]
  
  # Install Missing Custom Tools
  for (tool_name in missing_tools) {
    sub_dir <- tools[[tool_name]]
    pkg_path <- file.path(repo_root, sub_dir)
    
    # Fix DESCRIPTION file typos if present in the raw repo
    desc_file <- file.path(pkg_path, "DESCRIPTION")
    if (file.exists(desc_file)) {
      desc_lines <- readLines(desc_file)
      desc_lines <- gsub("Matrix,", "Matrix", desc_lines)
      writeLines(desc_lines, desc_file)
    }

    print(paste("🚀 Compiling", tool_name, "..."))
    remotes::install_local(pkg_path, dependencies = FALSE, force = TRUE, quiet = FALSE)
    library(tool_name, character.only = TRUE)
  }
  print("🎉 All tools installed and loaded successfully!")
}

# Run the installer check
install_cosmx_tools()
```


#Data Loading
```{r}
# ==============================================================================
# 2. MULTI-DATASET LOADING & INTEGRATION
# ==============================================================================

processed_matrices <- list()
processed_metadata <- list()

for (species in DATASET_TARGETS) {
  print(paste("=================================================="))
  print(paste("🔄 PROCESSING DATASET:", species))
  print(paste("=================================================="))
  
  # --- A. DYNAMIC PATH ROUTING ---
  if (species == "NHP") {
    data_dir <- file.path(BASE_DIR, "flatFiles", "S42NHPArray150")
    meta_path <- file.path(data_dir, "NHP CosMX TMA metadata.csv")
  } else if (species == "Human") {
    data_dir <- file.path(BASE_DIR, "flatFiles", "HumanTMA")
    meta_path <- file.path(data_dir, "Human CosMX TMA metadata.csv")
  }
  
  # Find the flat files using regex so exact file naming variations don't break the script
  expr_file <- list.files(data_dir, pattern = ".*_exprMat_file.csv.gz", full.names = TRUE)[1]
  base_meta_file <- list.files(data_dir, pattern = ".*_metadata_file.csv.gz", full.names = TRUE)[1]
  
  if (is.na(base_meta_file)) stop(paste("❌ Base metadata not found for", species, "in", data_dir))
  
  # --- B. METADATA PARSING & PREFIXING ---
  metadata <- fread(base_meta_file)
  
  # Merge custom FOV metadata if it exists
  if (file.exists(meta_path)) {
    fov_metadata <- fread(meta_path)
    if ("FOV" %in% colnames(fov_metadata)) setnames(fov_metadata, "FOV", "fov")
    metadata$fov <- as.integer(metadata$fov)
    metadata <- merge(metadata, fov_metadata, by = "fov", all.x = TRUE)
  }
  
  # Append species prefix to prevent cell_ID collisions between the arrays
  metadata$Species <- species
  metadata$cell_id_unique <- paste0(species, "_", metadata$fov, "_", metadata$cell_ID)
  rownames(metadata) <- metadata$cell_id_unique
  
  # --- C. DYNAMIC SUBSETTING (CASE-INSENSITIVE) ---
  metadata_subset <- metadata
  
  if (!is.null(TARGET_VALUES)) {
    target_col_lower <- tolower(SUBSET_COLUMN)
    cols_lower <- tolower(colnames(metadata))
    
    if (!(target_col_lower %in% cols_lower)) {
      print(paste("❌ ERROR: Column matching '", SUBSET_COLUMN, "' not found in", species, "metadata!"))
      print(paste("Available columns are:", paste(colnames(metadata), collapse=", ")))
      stop("Please update SUBSET_COLUMN in the User Configuration.")
    }
    
    actual_col_name <- colnames(metadata)[match(target_col_lower, cols_lower)]
    target_column_data <- metadata[[actual_col_name]]
    
    keep_rows <- grepl(paste(TARGET_VALUES, collapse="|"), target_column_data, ignore.case = TRUE)
    metadata_subset <- metadata[keep_rows, ]
    
    if (nrow(metadata_subset) == 0) {
      print(paste("⚠️ WARNING: No cells found for", species, "using targets:", paste(TARGET_VALUES, collapse=", ")))
      next 
    }
  }
  print(paste("✅ Retained", nrow(metadata_subset), "cells for", species))
  
  # --- D. MEMORY-OPTIMIZED MATRIX LOADING ---
  # Only load the count data for the cells that survived the subsetting
  counts_wide <- fread(expr_file)
  
  valid_fovs <- unique(metadata_subset$fov)
  counts_wide <- counts_wide[fov %in% valid_fovs]
  
  counts_wide[, cell_id_unique := paste0(species, "_", fov, "_", cell_ID)]
  counts_wide <- counts_wide[cell_id_unique %in% metadata_subset$cell_id_unique]
  
  meta_cols <- c("fov", "cell_ID", "cell_id_unique")
  gene_cols <- setdiff(colnames(counts_wide), meta_cols)
  
  mat_data <- as.matrix(counts_wide[, ..gene_cols])
  rownames(mat_data) <- counts_wide$cell_id_unique
  
  # Convert to a sparse matrix to save memory before merging
  mat <- Matrix::t(as(mat_data, "dgCMatrix"))
  rm(counts_wide, mat_data); gc()
  
  processed_matrices[[species]] <- mat
  processed_metadata[[species]] <- metadata_subset
}

# ==============================================================================
# 3. MERGE DATASETS & CREATE SEURAT OBJECT
# ==============================================================================
print("==================================================")
print("🧬 MERGING DATASETS AND INITIALIZING SEURAT")
print("==================================================")

if (length(processed_matrices) == 0) stop("❌ ERROR: No valid data loaded. Check your targets.")

# Combine the sparse matrices and metadata tables
final_mat <- do.call(cbind, processed_matrices)

final_meta <- rbindlist(processed_metadata, fill = TRUE)
final_meta <- as.data.frame(final_meta)
rownames(final_meta) <- final_meta$cell_id_unique

# Ensure metadata rows perfectly match the matrix columns
final_meta <- final_meta[match(colnames(final_mat), final_meta$cell_id_unique), ]

seurat_obj <- CreateSeuratObject(counts = final_mat, meta.data = final_meta, project = "Comparative_TMA")

print("🎉 SUCCESS: Merged Seurat Object Created.")
print(seurat_obj)
print(table(seurat_obj$Species))
```

#Quality Control
```{r}
# ==============================================================================
# QUALITY CONTROL & VISUALIZATION
# ==============================================================================
library(ggplot2)
library(patchwork)

print("Calculating QC Metrics...")

# Calculate Negative Probe Ratio (Background noise)
if ("nCount_negprobes" %in% colnames(seurat_obj@meta.data)) {
  seurat_obj$propNegative <- seurat_obj$nCount_negprobes / (seurat_obj$nCount_RNA + seurat_obj$nCount_negprobes)
} else {
  print("⚠️ 'nCount_negprobes' not found. Setting propNegative to 0.")
  seurat_obj$propNegative <- 0
}

# Generate QC Violin Plots grouped by Species
qc_features <- c("nFeature_RNA", "nCount_RNA", "propNegative", "Area")

# We group by Species to compare the arrays side-by-side
print(VlnPlot(seurat_obj, features = qc_features, group.by = "Species", ncol = 4, pt.size = 0))

# Feature Scatters to look for correlations
p1 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", group.by = "Species")
p2 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "propNegative", group.by = "Species")
print(p1 + p2)
```

#Metadata Download for LLM ingestion
```{r}
# ==============================================================================
# SPECIAL QC DIAGNOSTICS & EXPORT
# ==============================================================================
library(dplyr)

print("Generating Deep QC Plots...")

p1 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "propNegative", group.by = "Species") + 
  geom_hline(yintercept = 0.10, color = "red", linetype = "dashed") + 
  ggtitle("Negative Probes vs Counts")

p2 <- FeatureScatter(seurat_obj, feature1 = "Area", feature2 = "nFeature_RNA", group.by = "Species") + 
  geom_vline(xintercept = 20000, color = "red", linetype = "dashed") +
  ggtitle("Cell Area vs Gene Complexity")

# Spatial QC Check: Create a unique Species + FOV identifier to prevent cross-array merging
seurat_obj$Species_FOV <- paste0(seurat_obj$Species, "_FOV", seurat_obj$fov)

fov_qc <- seurat_obj@meta.data %>%
  group_by(Species_FOV) %>%
  summarize(
    Mean_Negatives = mean(propNegative),
    Mean_Counts = mean(nCount_RNA),
    Species = first(Species)
  )

p3 <- ggplot(fov_qc, aes(x = Mean_Counts, y = Mean_Negatives, label = Species_FOV, color = Species)) +
  geom_point() +
  geom_text(size = 3, vjust = -0.5, show.legend = FALSE) +
  ggtitle("Quality by FOV (Outlier Detection)") +
  theme_bw()

print((p1 + p2) / p3)

# EXPORT METADATA (Dynamic Naming)
meta_to_save <- seurat_obj@meta.data[, c("Species_FOV", "cell_ID", "nCount_RNA", "nFeature_RNA", "Area", "propNegative")]
export_name <- paste0("Comparative_QC_Metadata_", Sys.Date(), ".csv")
write.csv(meta_to_save, file = file.path(BASE_DIR, export_name))
print(paste("✅ Metadata saved to:", file.path(BASE_DIR, export_name)))

# ==============================================================================
# ARTIFACT CHARACTERIZATION
# ==============================================================================
print("Analyzing False Positives in 'Artifact' Regions...")

if ("artifact" %in% colnames(seurat_obj@meta.data)) {
  seurat_obj$is_artifact <- seurat_obj$artifact == TRUE
  n_artifact <- sum(seurat_obj$is_artifact, na.rm = TRUE)
  
  print(paste("Found", n_artifact, "cells in regions labeled 'Artifact'."))

  if (n_artifact > 0) {
    ap1 <- VlnPlot(seurat_obj, features = "nCount_RNA", group.by = "is_artifact", pt.size = 0, log = TRUE) + NoLegend()
    ap2 <- VlnPlot(seurat_obj, features = "propNegative", group.by = "is_artifact", pt.size = 0) + NoLegend()
    print(ap1 + ap2)
    
    ghost_cells <- seurat_obj@meta.data[seurat_obj$is_artifact == TRUE, ]
    suggested_min_counts <- quantile(ghost_cells$nCount_RNA, 0.99)
    
    print("----------------------------------------------------------------")
    print("👻 ARTIFACT CELL STATISTICS (Background Noise Profile)")
    print(paste("Max Counts (99th percentile):   ", round(suggested_min_counts, 0)))
    print("----------------------------------------------------------------")
  }
} else {
  print("⚠️ Column 'artifact' not found. Skipping artifact noise characterization.")
}
```

#Filtering
```{r}
# ==============================================================================
# BATCH-AWARE QUALITY CONTROL & FILTERING
# ==============================================================================
print("Starting Batch-Aware QC...")

# USER CONFIG: Define which column represents your batches (e.g., "Species", "AnimalID", "Patient")
BATCH_COL <- "Species" 

if (!(BATCH_COL %in% colnames(seurat_obj@meta.data))) stop(paste("❌ BATCH_COL", BATCH_COL, "not found!"))

# 1. VISUALIZE BATCH EFFECTS
p1 <- VlnPlot(seurat_obj, features = "nCount_RNA", group.by = BATCH_COL, pt.size = 0, log = TRUE) + 
  geom_hline(yintercept = 50, color = "red", linetype = "dashed")
p2 <- VlnPlot(seurat_obj, features = "propNegative", group.by = BATCH_COL, pt.size = 0) + 
  geom_hline(yintercept = 0.05, color = "red", linetype = "dashed")
print(p1 + p2)

# 2. CALCULATE BATCH STATISTICS
batch_stats <- seurat_obj@meta.data %>%
  group_by(!!sym(BATCH_COL)) %>%
  summarize(
    Median_Counts = median(nCount_RNA),
    Median_Noise = median(propNegative),
    Cell_Count = n()
  )
print("--- Batch Statistics ---")
print(batch_stats)

# 3. APPLY "FORGIVING" FILTERS
MIN_COUNTS <- 50 
MAX_NEGATIVE <- 0.10 
MAX_AREA <- 20000

print(paste("Applying Filters: Counts >", MIN_COUNTS, "| Neg <", MAX_NEGATIVE*100, "% | Area <", MAX_AREA))

seurat_subset <- subset(seurat_obj, 
                        subset = nFeature_RNA > 20 & 
                                 nCount_RNA > MIN_COUNTS & 
                                 propNegative < MAX_NEGATIVE & 
                                 Area < MAX_AREA)

cells_before <- ncol(seurat_obj)
cells_after <- ncol(seurat_subset)
print(paste("Original Cells:", cells_before, "| Remaining:", cells_after, "| Removed:", cells_before - cells_after))

rm(seurat_obj); gc()
```

#Normalization and Dimensionality Reduction
```{r}
# ==============================================================================
# NORMALIZATION & DIMENSIONALITY REDUCTION (scPearsonPCA)
# ==============================================================================
print("Running scPearsonPCA Normalization...")

tc <- Matrix::colSums(seurat_subset[["RNA"]]$counts) 
genefreq <- scPearsonPCA::gene_frequency(seurat_subset[["RNA"]]$counts) 

seurat_subset <- FindVariableFeatures(seurat_subset, selection.method = "vst", nfeatures = 2000)
hvgs <- VariableFeatures(seurat_subset)

set.seed(42) 

pcaobj <- sparse_quasipoisson_pca_seurat(
  seurat_subset[["RNA"]]$counts[hvgs,],
  totalcounts = tc, 
  grate = genefreq[hvgs],
  scale.max = 10, 
  do.scale = TRUE, 
  do.center = TRUE
)

seurat_subset[["pca"]] <- pcaobj$reduction.data
print("✅ Normalization & PCA Complete.")

print(ElbowPlot(seurat_subset, ndims = 50))
gc()
```

#UMAP & Clustering
```{r}
# ==============================================================================
# BATCH INTEGRATION (HARMONY) & CLUSTERING
# ==============================================================================
if (!require("harmony", quietly = TRUE)) install.packages("harmony")
library(harmony)
library(patchwork)

print("Running Harmony Integration & Clustering...")

# USER CONFIG: What metadata column defines your batches for integration?
INTEGRATION_VAR <- "Species" 
SEED_VALUE <- 42
set.seed(SEED_VALUE)

# 0. Generate Uncorrected UMAP (Before Harmony)
# We save this to a custom reduction name so it doesn't get overwritten
print("Generating uncorrected UMAP for baseline comparison...")
seurat_subset <- RunUMAP(seurat_subset, reduction = "pca", dims = 1:20, 
                         reduction.name = "umap_uncorrected", 
                         reduction.key = "UMAPuncorrected_",
                         seed.use = SEED_VALUE)

# 1. Run Harmony (Integrate across Species arrays)
print("Applying Harmony Batch Correction...")
seurat_subset <- RunHarmony(object = seurat_subset, 
                            group.by.vars = INTEGRATION_VAR, 
                            reduction.use = "pca",        
                            assay.use = "RNA", 
                            project.dim = FALSE,
                            plot_convergence = TRUE)

# 2. Run Corrected UMAP & Find Neighbors
print("Generating corrected UMAP and Clusters...")
seurat_subset <- RunUMAP(seurat_subset, reduction = "harmony", dims = 1:20, 
                         reduction.name = "umap", 
                         reduction.key = "UMAP_",
                         seed.use = SEED_VALUE)
seurat_subset <- FindNeighbors(seurat_subset, reduction = "harmony", dims = 1:20)
seurat_subset <- FindClusters(seurat_subset, resolution = 0.3, random.seed = SEED_VALUE)

# 3. Visualization: Before vs After Batch Correction
p_before <- DimPlot(seurat_subset, reduction = "umap_uncorrected", group.by = INTEGRATION_VAR) + 
  ggtitle(paste("BEFORE Harmony:", INTEGRATION_VAR)) +
  theme(legend.position = "bottom")

p_after <- DimPlot(seurat_subset, reduction = "umap", group.by = INTEGRATION_VAR) + 
  ggtitle(paste("AFTER Harmony:", INTEGRATION_VAR)) +
  theme(legend.position = "bottom")

print("Visualizing Batch Correction Effect:")
print(p_before | p_after)

# 4. Visualization: Biological Clusters & Subset Check
p_clusters <- DimPlot(seurat_subset, reduction = "umap", label = TRUE) + 
  ggtitle("Clusters (Cell Types)")

# Dynamically find the exact case-sensitive column name to prevent 'undefined columns' errors
actual_plot_col <- grep(paste0("^", SUBSET_COLUMN, "$"), colnames(seurat_subset@meta.data), ignore.case = TRUE, value = TRUE)

p_subset <- DimPlot(seurat_subset, reduction = "umap", group.by = actual_plot_col) + 
  ggtitle(paste("Subset Check:", actual_plot_col))

print("Visualizing Downstream Biology:")
print(p_clusters | p_subset)
```


#To Cluster or Not to Cluster
```{r}
#This block 1) Test multiple clustering resolutions 2) Checks for new clusters and the markers that define them 3) Checks cell size of new clusters (to make sure they aren't too small for downstream analysis) 4) Sets a winning resolution for future blocks

#The purpose of this block is to determine the best resolution for clustering. The UMAPs in the block above this are at a resolution of 0.3, but you don't know if thats overclustering or underclustering (its just a starting point to see if you will get clusters). This block tests resolutions from (0.3-1, but you can change them). At each resolution level it generates a list of markers defining the cluster. This helps you decide if a new cluster is formed, is it a cluster that makes biological sense or is it a nonsensical split.

#When complete you should check the .csv files this exports to analyze the markers definine the new clusters and to determine which cell types likely correspond to each of those clusters. !! In some instances this may not be appropriate because a diseased cell cluster may not express cannonical markers !!

# ==============================================================================
# MULTI-RESOLUTION SWEEP & MARKER DISCOVERY
# ==============================================================================
library(clustree)

resolutions_to_test <- c(0.3, 0.5, 0.8, 1.0)
print("🚀 Starting Multi-Resolution Sweep...")

seurat_subset <- FindClusters(seurat_subset, resolution = resolutions_to_test, random.seed = 42, verbose = FALSE)
print(clustree(seurat_subset, prefix = "RNA_snn_res.") + ggtitle("Cluster Stability Tree"))

# USER CONFIG: Change this to the winning resolution based on the Clustree
winner_res <- 0.5 
final_markers <- NULL

print("Applying Log-Normalization for Marker Search...")
seurat_subset <- NormalizeData(seurat_subset, normalization.method = "LogNormalize", scale.factor = 10000)

for (res in resolutions_to_test) {
  Idents(seurat_subset) <- paste0("RNA_snn_res.", res)
  print(paste("🔎 ANALYZING RESOLUTION:", res))
  
  markers <- FindAllMarkers(seurat_subset, only.pos = TRUE, min.pct = 0.1, logfc.threshold = 0.25, verbose = FALSE)
  
  # Save to dynamic file path
  file_name <- paste0("MultiSpecies_Markers_Res_", res, ".csv")
  write.csv(markers, file = file.path(BASE_DIR, file_name))
  
  if (res == winner_res) {
    final_markers <- markers
    print(paste("✅ Resolution", res, "selected as Final. Markers loaded into memory."))
  }
}

Idents(seurat_subset) <- paste0("RNA_snn_res.", winner_res)

top_markers <- final_markers %>%
  group_by(cluster) %>%
  slice_max(n = 20, order_by = avg_log2FC)

write.csv(top_markers, file = file.path(BASE_DIR, "MultiSpecies_Top_Markers_Final.csv"))
print("✅ Optimized Sweep Complete. Ready for Annotation.")
```

#Assign Cell Types/Names to each Cluster
```{r}
#Note this entire block is dependent on the results/interpretation of the prior block and the resolution you ultimately select. The MultiSpecies_Markers for the chosen resolution can be passed to the LLM to assist with initial cluster identifications/naming

# ==============================================================================
# STEP 8: HIERARCHICAL ANNOTATION
# ==============================================================================
print("Starting Cell Annotation...")

# 1. SET THE GLOBAL BASELINE (Resolution 0.5)
Idents(seurat_subset) <- "RNA_snn_res.0.5"

# 2. RENAME THE CLUSTERS BASED ON BIOLOGY
# The numeric IDs below match the markers found in the Res 0.5 CSV.
base_cluster_ids <- c(
  "0" = "Hepatocytes (CRHR2+)",      # Specialized/Altered subset
  "1" = "Immune (Kupffer/Macs)",       # C1QB, HLA-DRA
  "2" = "Hepatocytes (Canonical)",   # SERPINA1, TTR, FGG
  "3" = "Endothelial (LSECs)",         # LDB2, KDR
  "4" = "Fibroblasts / Portal Stroma", # DCN, CXCL12, Collagen
  "5" = "Stellate / Smooth Muscle"     # ACTA2, TAGLN, MYH11
)

seurat_subset <- RenameIdents(seurat_subset, base_cluster_ids)

# We save this final identity into the metadata for all downstream DEGs
seurat_subset$CellType_Final <- Idents(seurat_subset)
print("✅ High-Resolution Cell Types Assigned.")

# ------------------------------------------------------------------------------
# 3. VISUALIZATION
# ------------------------------------------------------------------------------
library(patchwork)

# Generate the labeled UMAP (Explicitly forcing the Harmony-corrected map)
p_final <- DimPlot(seurat_subset, 
                   reduction = "umap",           # <-- THE FIX
                   group.by = "CellType_Final", 
                   label = TRUE, 
                   repel = TRUE) + 
  ggtitle("Final Annotated Cell Types (Res 0.5)") + 
  theme(legend.position = "right")

print(p_final)
```

#DEG
```{r}
#Changing the DE_GROUP_COL will switch these between Species or CellTypes. The Species comparison is primarily what we want to check (differences between NHPs and Humans). The CellType check, helps you better understand your clustering and decide if you need additional clustering or not

# ==============================================================================
# STEP 9: DIFFERENTIAL EXPRESSION & VOLCANO PLOTS
# ==============================================================================
library(ggplot2)
library(ggrepel)
library(dplyr)

# ------------------------------------------------------------------------------
# USER CONFIGURATION FOR DIFFERENTIAL EXPRESSION
# ------------------------------------------------------------------------------
# Define the metadata column to split your data by (e.g., "Species", "CellType_Final")
DE_GROUP_COL <- "Species"

# Define the two groups you want to compare within that column. 
# Positive Log2FC will mean the gene is higher in GROUP_1.
DE_GROUP_1 <- "Human"  
DE_GROUP_2 <- "NHP"    

print(paste("Running Differential Expression:", DE_GROUP_1, "vs", DE_GROUP_2))

# ------------------------------------------------------------------------------
# HELPER FUNCTION FOR VOLCANO PLOTS
# ------------------------------------------------------------------------------
plot_volcano <- function(de_results, title_text, group1_name, group2_name) {
  de_results$diffexpressed <- "NO"
  de_results$diffexpressed[de_results$avg_log2FC > 0.5 & de_results$p_val_adj < 0.05] <- paste("Higher in", group1_name)
  de_results$diffexpressed[de_results$avg_log2FC < -0.5 & de_results$p_val_adj < 0.05] <- paste("Higher in", group2_name)
  
  de_results$delabel <- NA
  top_genes <- head(de_results[order(de_results$p_val_adj), "Gene"], 15)
  de_results$delabel[de_results$Gene %in% top_genes] <- de_results$Gene[de_results$Gene %in% top_genes]
  
  # Set dynamic colors
  custom_colors <- c("grey", "red", "blue")
  names(custom_colors) <- c("NO", paste("Higher in", group1_name), paste("Higher in", group2_name))
  
  p <- ggplot(data = de_results, aes(x = avg_log2FC, y = -log10(p_val_adj), col = diffexpressed, label = delabel)) +
    geom_point(alpha = 0.7) + 
    theme_minimal() +
    geom_text_repel(max.overlaps = 15, show.legend = FALSE) +
    scale_color_manual(values = custom_colors) +
    geom_vline(xintercept = c(-0.5, 0.5), col = "red", linetype="dashed", alpha = 0.3) +
    geom_hline(yintercept = -log10(0.05), col = "red", linetype="dashed", alpha = 0.3) +
    ggtitle(title_text) +
    labs(x = paste("Log2 Fold Change (", group1_name, "/", group2_name, ")"), 
         y = "-Log10 Adjusted P-value",
         color = "Significance") +
    theme(legend.position = "bottom")
  
  return(p)
}

# ------------------------------------------------------------------------------
# PART A: GLOBAL DIFFERENTIAL EXPRESSION
# ------------------------------------------------------------------------------
print("🌍 Running Global Analysis...")

# Create a safe file name component
safe_g1 <- gsub("[^A-Za-z0-9]", "_", DE_GROUP_1)
safe_g2 <- gsub("[^A-Za-z0-9]", "_", DE_GROUP_2)

tryCatch({
  global_markers <- FindMarkers(seurat_subset, 
                                ident.1 = DE_GROUP_1, 
                                ident.2 = DE_GROUP_2, 
                                group.by = DE_GROUP_COL,
                                min.pct = 0.1, 
                                logfc.threshold = 0.1)
  
  global_markers$Gene <- rownames(global_markers)
  write.csv(global_markers, file = file.path(BASE_DIR, paste0("DE_", safe_g1, "_vs_", safe_g2, ".csv")))
  
  p_global <- plot_volcano(global_markers, paste(DE_GROUP_1, "vs", DE_GROUP_2), DE_GROUP_1, DE_GROUP_2)
  print(p_global)
  ggsave(filename = file.path(BASE_DIR, paste0("Volcano_", safe_g1, "_vs_", safe_g2, ".png")), plot = p_global, width = 7, height = 6)
  print("✅ Global Analysis Complete.")
  
}, error = function(e) { print(paste("⚠️ Global DE Failed:", e$message)) })

# ------------------------------------------------------------------------------
# PART B: PER-CLUSTER DIFFERENTIAL EXPRESSION (Only if comparing conditions)
# ------------------------------------------------------------------------------
# If the user is comparing two Cell Types against each other, doing a per-cluster 
# breakdown makes no sense (you can't compare Hepatocytes vs CRHR2 within Hepatocytes).
# We only run this block if DE_GROUP_COL is something broad like "Species" or "Malaria".

if (DE_GROUP_COL != "CellType_Final" && DE_GROUP_COL != "CellType_Global") {
  
  Idents(seurat_subset) <- "CellType_Final" 
  clusters_to_test <- unique(Idents(seurat_subset))
  
  print("🔍 Running DE Analysis within each Cell Type...")
  results_list <- list()
  
  for (cell_type in clusters_to_test) {
    safe_name <- gsub("[^A-Za-z0-9]", "_", cell_type)
    cluster_cells <- subset(seurat_subset, idents = cell_type)
    
    n_group1 <- sum(cluster_cells[[DE_GROUP_COL]] == DE_GROUP_1)
    n_group2 <- sum(cluster_cells[[DE_GROUP_COL]] == DE_GROUP_2)
    
    if (n_group1 > 10 & n_group2 > 10) {
      print(paste("   Testing:", cell_type, "|", DE_GROUP_1, ":", n_group1, "vs", DE_GROUP_2, ":", n_group2))
      
      tryCatch({
        de_markers <- FindMarkers(cluster_cells, 
                                  ident.1 = DE_GROUP_1, 
                                  ident.2 = DE_GROUP_2, 
                                  group.by = DE_GROUP_COL,
                                  min.pct = 0.1,       
                                  logfc.threshold = 0.1,
                                  verbose = FALSE) 
        
        if (nrow(de_markers) > 0) {
          de_markers$Cluster <- cell_type
          de_markers$Gene <- rownames(de_markers)
          results_list[[as.character(cell_type)]] <- de_markers
          
          p_cluster <- plot_volcano(de_markers, paste(cell_type, ":", DE_GROUP_1, "vs", DE_GROUP_2), DE_GROUP_1, DE_GROUP_2)
          print(p_cluster) 
          ggsave(filename = file.path(BASE_DIR, paste0("Volcano_", safe_name, "_", safe_g1, "_vs_", safe_g2, ".png")), plot = p_cluster, width = 7, height = 6)
        }
      }, error = function(e) {
        print(paste("   ⚠️ Error in", cell_type, ":", e$message))
      })
      
    } else {
      print(paste("   ⚠️ Skipping", cell_type, "- Insufficient cells in one or both groups."))
    }
  }
  
  if (length(results_list) > 0) {
    deg_table <- do.call(rbind, results_list)
    write.csv(deg_table, file = file.path(BASE_DIR, paste0("DE_By_CellType_", safe_g1, "_vs_", safe_g2, ".csv")))
    print("✅ Per-Cluster Analysis Complete.")
  } else {
    print("⚠️ No DE genes found in any cluster.")
  }

} else {
  print("ℹ️ Skipping Per-Cluster Analysis because groups are Cell Types.")
}
```


####STOP LST MEETING#####

#DEG for Specific Genes
```{r}
#This block is for DEG of specific genes of interest. If you wanted to curate a list of TH1 related genes to see how they compare. Note that DEG may not be the best for this and instead you may want to generate an expression heatmap.

# ==============================================================================
# STEP 10: TARGETED GENE PANEL VOLCANO PLOT
# ==============================================================================
# Goal: Visualize the global shift in a specific, user-defined set of genes.

library(ggplot2)
library(ggrepel)

print("🚀 Running Targeted DEG for Volcano Plot...")

# ------------------------------------------------------------------------------
# 1. DEFINE YOUR GENE PANEL
# ------------------------------------------------------------------------------
# Fill this list with your genes of interest. 
# Example: target_panel <- c("ACTA2", "COL1A1", "TGFB1")
target_panel <- c(
  # Add genes here
)

if(length(target_panel) == 0) {
  print("⚠️ Target panel is empty. Please add genes to the 'target_panel' list to run this block.")
} else {
  
  # Keep only genes present in the dataset
  target_panel <- intersect(target_panel, rownames(seurat_subset))
  print(paste("Found", length(target_panel), "target genes in the dataset."))
  
  # 2. Run Targeted Differential Expression
  target_deg <- FindMarkers(seurat_subset, 
                            ident.1 = DE_GROUP_1, 
                            ident.2 = DE_GROUP_2, 
                            group.by = DE_GROUP_COL, 
                            features = target_panel, 
                            min.pct = 0,         # Force inclusion of rare genes
                            logfc.threshold = 0, # Force inclusion of subtle changes
                            verbose = FALSE)
  
  # 3. Prepare Data for Plotting
  target_deg$Gene <- rownames(target_deg)
  target_deg$Significance <- "NS"
  target_deg$Significance[target_deg$p_val < 0.05 & target_deg$avg_log2FC > 0] <- paste("Up in", DE_GROUP_1)
  target_deg$Significance[target_deg$p_val < 0.05 & target_deg$avg_log2FC < 0] <- paste("Up in", DE_GROUP_2)
  
  # Define custom colors
  custom_colors <- c("blue", "red", "grey")
  names(custom_colors) <- c(paste("Up in", DE_GROUP_2), paste("Up in", DE_GROUP_1), "NS")
  
  # 4. Generate the Volcano Plot
  p_target_volcano <- ggplot(target_deg, aes(x = avg_log2FC, y = -log10(p_val), color = Significance, label = Gene)) +
    geom_point(size = 3, alpha = 0.8) +
    geom_text_repel(max.overlaps = 20, box.padding = 0.5, show.legend = FALSE) + 
    scale_color_manual(values = custom_colors) +
    geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.5) +
    theme_bw() +
    ggtitle(paste("Targeted Panel Shift:", DE_GROUP_1, "vs", DE_GROUP_2)) +
    xlab(paste("Log2 Fold Change (Positive =", DE_GROUP_1, ")")) +
    ylab("-Log10 P-Value") +
    theme(legend.position = "bottom")
  
  # 5. Display and Save
  print(p_target_volcano)
  ggsave(file.path(BASE_DIR, paste0("Targeted_Volcano_", DE_GROUP_1, "_vs_", DE_GROUP_2, ".png")), plot = p_target_volcano, width = 8, height = 6)
  
  # Print the Table
  print(paste("--- TARGETED SHIFTS:", DE_GROUP_1, "vs", DE_GROUP_2, "---"))
  print(target_deg[order(target_deg$avg_log2FC), c("avg_log2FC", "p_val")])
}
```

#Common and Differential Expression between clusters
```{r}
#Not working right now. Built to further check UMAP clustering in a detailed manner.

# ==============================================================================
# STEP 9.5: LINEAGE RELATIONSHIP ANALYSIS (CRHR2+ vs Hepatocytes)
# ==============================================================================
library(ggplot2)
library(ggrepel)
library(dplyr)

print("🚀 Running Lineage Relationship Analysis...")

# ------------------------------------------------------------------------------
# 1. USER CONFIGURATION
# ------------------------------------------------------------------------------
# Define the two clusters you suspect are related
GROUP_A <- "Hepatocytes"
GROUP_B <- "Hepatocytes (High-RNA / Altered)"
IDENT_COL <- "CellType_Final"

Idents(seurat_subset) <- IDENT_COL

# ------------------------------------------------------------------------------
# 2. FIND SHARED LINEAGE MARKERS (Group A + Group B vs Everyone Else)
# ------------------------------------------------------------------------------
print("🔍 Identifying shared lineage markers...")
# Temporarily group them together to see what makes them BOTH unique from the rest of the liver
seurat_subset$Temp_Lineage <- as.character(Idents(seurat_subset))
seurat_subset$Temp_Lineage[seurat_subset$Temp_Lineage %in% c(GROUP_A, GROUP_B)] <- "Target_Lineage"

Idents(seurat_subset) <- "Temp_Lineage"
shared_markers <- FindMarkers(seurat_subset, 
                              ident.1 = "Target_Lineage", 
                              only.pos = TRUE, 
                              min.pct = 0.25, 
                              logfc.threshold = 0.5,
                              verbose = FALSE)

top_shared <- head(rownames(shared_markers), 8) # Grab top 8 shared genes
print(paste("Top Shared Markers:", paste(top_shared, collapse = ", ")))

# ------------------------------------------------------------------------------
# 3. FIND DIFFERENTIATING MARKERS (Group A vs Group B)
# ------------------------------------------------------------------------------
print("⚔️ Identifying differentiating state markers...")
Idents(seurat_subset) <- IDENT_COL # Reset back to normal cell types

direct_deg <- FindMarkers(seurat_subset, 
                          ident.1 = GROUP_A, 
                          ident.2 = GROUP_B, 
                          verbose = FALSE)

# Get top 5 unique to A and top 5 unique to B
top_unique_A <- head(rownames(direct_deg[direct_deg$avg_log2FC > 0 & direct_deg$p_val_adj < 0.05, ]), 5)
top_unique_B <- head(rownames(direct_deg[direct_deg$avg_log2FC < 0 & direct_deg$p_val_adj < 0.05, ]), 5)

# ------------------------------------------------------------------------------
# 4. AVERAGE EXPRESSION SCATTER PLOT
# ------------------------------------------------------------------------------
print("📊 Generating Co-Expression Scatter Plot...")

# Calculate average expression per cluster
avg_exp <- AverageExpression(seurat_subset, group.by = IDENT_COL, assays = "RNA")$RNA

# Create a dataframe for plotting
df_scatter <- data.frame(
  Gene = rownames(avg_exp),
  GroupA_Exp = log1p(avg_exp[, GROUP_A]), # log1p smooths the extreme spatial counts
  GroupB_Exp = log1p(avg_exp[, GROUP_B])
)

# Label the genes we found in steps 2 and 3
df_scatter$GeneType <- "Other"
df_scatter$GeneType[df_scatter$Gene %in% top_shared] <- "Shared Lineage"
df_scatter$GeneType[df_scatter$Gene %in% top_unique_A] <- paste("Unique to", GROUP_A)
df_scatter$GeneType[df_scatter$Gene %in% top_unique_B] <- paste("Unique to", GROUP_B)

# Set up specific colors
scatter_colors <- c("grey80", "purple", "blue", "red")
names(scatter_colors) <- c("Other", "Shared Lineage", paste("Unique to", GROUP_A), paste("Unique to", GROUP_B))

p_scatter <- ggplot(df_scatter, aes(x = GroupA_Exp, y = GroupB_Exp, color = GeneType)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black", alpha = 0.5) +
  geom_text_repel(data = subset(df_scatter, GeneType != "Other"), aes(label = Gene), 
                  max.overlaps = 30, show.legend = FALSE, size = 4, fontface = "bold") +
  scale_color_manual(values = scatter_colors) +
  theme_bw() +
  ggtitle(paste("Lineage Comparison:", GROUP_A, "vs", GROUP_B)) +
  xlab(paste("Log1p Average Expression -", GROUP_A)) +
  ylab(paste("Log1p Average Expression -", GROUP_B)) +
  theme(legend.position = "bottom", legend.title = element_blank())

print(p_scatter)
ggsave(filename = file.path(BASE_DIR, "Lineage_Scatter_Plot.png"), plot = p_scatter, width = 7, height = 7)

# ------------------------------------------------------------------------------
# 5. COMPARATIVE DOT PLOT
# ------------------------------------------------------------------------------

print("🟢 Generating Comparative DotPlot...")

# Combine the genes into one list for the DotPlot
features_to_plot <- c(top_shared, top_unique_A, top_unique_B)

p_dot <- DotPlot(seurat_subset, features = features_to_plot, group.by = IDENT_COL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Marker Validation Across All Clusters") +
  xlab("Genes (Shared -> Unique A -> Unique B)") +
  ylab("Cell Type")

print(p_dot)
ggsave(filename = file.path(BASE_DIR, "Lineage_DotPlot.png"), plot = p_dot, width = 10, height = 5)

print("✅ Lineage Analysis Complete.")
```

